Naturalness in U(1) extensions of the MSSM

### 2015-01-15 paper/e6ssm-tuning-paper.tex
Added more neutralino discussion.

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}
\usepackage{hyperref}
\usepackage{footmisc}
% macros
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{eqnarray}}
\newcommand{\ea}{\end{eqnarray}}
\newcommand{\SuperField}[1]{\hat{#1}}

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
for two different values of the $Z^\prime$ mass. We choose to look at
$M_{Z^\prime}=2.5$ TeV, which is just above the current limits, and
$M_{Z^\prime}=4.5$ TeV, which should be in reach in run II at the LHC
and then compare the fine tuning calculated in each case to the tuning
in the MSSM.  For this, we have performed a large six dimensional
parameter space scan in both the MSSM and E$_6$SSM, varying those
parameters most relevant for the fine tuning and the Higgs mass.
Therefore the set of parameters which we vary includes $\mu$, $B$ and
$\tan \beta$ for the MSSM, and $\lambda$, $A_\lambda$ and $\tan
\beta$, for the E$_6$SSM, which appear at tree level in the EWSB
conditions of the models. While the contribution from large stop
masses to the fine tuning is small for such a low cut-off scale, the
stop contributions to the effective potential can play a significant
role in reducing the fine tuning.  For this reason it is still
important to properly treat the tuning associated with stop
contributions to the one-loop effective potential, and so we also scan
over the soft masses $m_{Q_3}^2$, $m_{u_3}^2$ and the stop mixing
$A_t$.  The relevant parameters and ranges that were scanned over are
summarised in Table \ref{tab:scanranges}. With the exception of the
stop mixing $A_t$ and the soft trilinear $A_\lambda$, which are
logarithmically scanned, all parameters are linearly scanned.  In
addition to this we also repeat each scan for three different values
of $M_2$ to introduce more variation in the chargino masses, which can
also be importnat due to the link between $\mu$ and the fine tuning.

@@

%   For this we also fix [what we fix + explain different choices to
%   before and their impact].
\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{all_models_tuning_vs_higgs.png}}
\caption{Scatter plot of fine tuning vs lightest Higgs mass for the
  MSSM (green), E$_6$SSM with $M_{Z^\prime} = 2.5$ TeV (blue) and
  E$_6$SSM with $M_{Z^\prime} = 4.5$ TeV (red).}
\label{Fig:e6ssmvsmssm}
\end{center}
\end{figure}
In Fig \ref{Fig:e6ssmvsmssm} results from the scan are plotted showing
the tuning for each case against the lightest Higgs mass.  As expected
the dependence on the Higgs mass is now quite weak, while the minimum
tuning in the model for the E$_6$SSM is increased by the mass of the
$Z^\prime$ boson.  So in the case of a very low cutoff much of the
tuning required to get a 125 GeV Higgs has gone away.  However the
tuning from the $Z^\prime$ mass appears already at tree level and is
unavoidable regardless of the cutoff scale chosen, so in the E$_6$SSM
current limits on the mass of the $Z^\prime$ boson imply that the
E$_6$SSM fine tuning must be at least 100 regardless of assumptions
about the way SUSY breaking happens.  Additionally if run II of the
LHC pushes the limit on the $Z^\prime$ mass to be above $4.5$ TeV then
the fine tuning in the model will be greater than at least 350.

This demopnstrates two important points about these $U(1)$
extensions. Firstly that limits on the $Z^\prime$ mass play an
incredibly important role in constraining natural scenarios in such
models and secondly that the tuning from the $Z^\prime$ limits in
these models depends on less assumptions about SUSY breaking than the
tuning reqyired by the $125$ GeV Higgs measurement which concerns
people in the MSSM.

However we should also acknowledge that there are other limits which
play a similar role.  Chargino limits directly constrain the $\mu$
parameter (or effective $\mu$ parameter in these $U(1)$ extensions).
Unfortunately the chargino limits from the LHC are not general enough
so as to provide an absolute constraint on the model [Add: XSdetails on
  limits and what they depend on].

Nonetheless the MSSM the impact of chargino limit is shown in
Fig.~\ref{}.  There we see that if the the full parameter space with
$m_{\chi^\pm} < 700$ GeV was excluded the impact would be to make the
tuning in the MSSM with a $20$ TeV cutoff slightly worse than that of
the E$_6$SSM with the same cutoff and a $Z^\prime$ mass just larger
than current limits.  would be similar to that can also be attcked
more directly by looking at chargino mass limits as this rasies $\mu$
directly.  In the E$_6$SSM, while raising the chargino limit can have
the same impact in princuiple, due to current limits on the $Z^\prime$
mass already imposing a significant degree of tuning, chargino masses
do not make much of a noticable change.

[show plots in MSSM and E6SSM if it gets clened up, though maybe we
  can just leave it out if it only shows that there is no big impact?]
%    - plots of chargino bound impacts

The exact level of tuning from the $Z^\prime$ depends on the charges
of the extra $U(1)$ gauge symmetry it is associated with.  In
Fig.~\ref{Fig:othere6modelsvsmssm} we look at the fine tuning for
other U(1) extensions for the same $Z^\prime$ masses as we did for the
E$_6$SSM.  To simplify the analysis we fix $\tan \beta = 10$, but scan
over the remaining parameters as in Table \ref{tab:scanranges} and fix
the rest to the same values we did in the scan carried out for
Fig.~\ref{Fig:e6ssmvsmssm}
% - plots for different U(1)' charge

% DH:: for reference:
%      U(1)_I \theta = \pi+\arctan\sqrt{3/5}
%      U(1)_\psi \theta = \pi / 2
%      U(1)_\eta \theta = -\arctan\sqrt{5/3}

\begin{figure}[h]
\begin{center}
\resizebox{!}{5cm}{%
\includegraphics{e6inert_tb10_tuning_vs_higgs.png}}
\resizebox{!}{5cm}{%
\includegraphics{e6psi_tb10_tuning_vs_higgs.png}}\\
\resizebox{!}{5cm}{%
\includegraphics{e6eta_tb10_tuning_vs_higgs.png}}
\caption{Scatter plot of fine tuning vs lightest Higgs mass in ... with $M_{Z^\prime} = 2.5$ TeV shown in blue and $M_{Z^\prime} = 4.5$ TeV in red.}
\label{Fig:othere6modelsvsmssm}
\end{center}
\end{figure}

@@

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models, explaining why
% DH:: I strongly suspect that they are invalid - but running scan
% to check this currently
the heavier Higgs values are less well populated [can't be obtained? -
are we sure those scattered points are valid?]  than in the other
models.  Therefore the fine tuining behaviour in this model is closer
to that of the MSSM, though the LHC can still constrain the most
natural scenarions from just placing tougher limits on the $Z^\prime$
mass .

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}
\usepackage{hyperref}
\usepackage{footmisc}
% macros
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\ba}{\begin{eqnarray}}
\newcommand{\ea}{\end{eqnarray}}
\newcommand{\SuperField}[1]{\hat{#1}}
\usepackage{url}

@@

Since the stop mass does not have such a large impact on the fine
tuning when the cutoff scale is very low we can use this to see more
clearly the impact of the $Z^\prime$ mass on fine tuning. To do so we
select a fixed low cutoff of $M_X=20$ TeV and compare the fine tuning
for two different values of the $Z^\prime$ mass. We choose to look at
$M_{Z^\prime}=2.5$ TeV, which is just above the current limits, and
$M_{Z^\prime}=4.5$ TeV, which should be in reach in run II at the LHC
and then compare the fine tuning calculated in each case to the tuning
in the MSSM.  For this, we have performed a large six dimensional
parameter space scan in both the MSSM and E$_6$SSM, varying those
parameters most relevant for the fine tuning and the Higgs mass.
Therefore the set of parameters which we vary includes $\mu$, $B$ and
$\tan \beta$ for the MSSM, and $\lambda$, $A_\lambda$ and $\tan
\beta$, for the E$_6$SSM, which appear at tree level in the EWSB
conditions of the models. While the contribution from large stop
masses to the fine tuning is small for such a low cut-off scale, the
stop contributions to the effective potential can play a significant
role in reducing the fine tuning.  For this reason it is still
important to properly treat the tuning associated with stop
contributions to the one-loop effective potential, and so we also scan
over the soft masses $m_{Q_3}^2$, $m_{u_3}^2$ and the stop mixing
$A_t$.  The relevant parameters and ranges that were scanned over are
summarised in Table \ref{tab:scanranges}. With the exception of the
stop mixing $A_t$ and the soft trilinear $A_\lambda$, which are
logarithmically scanned, all parameters are linearly scanned.  In
addition to this we also repeat each scan for three different values
of $M_2$ to allow more variation in the chargino masses.

@@

%   For this we also fix [what we fix + explain different choices to
%   before and their impact].
\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{all_models_tuning_vs_higgs.png}}
\caption{Scatter plot of fine tuning vs lightest Higgs mass for the
  MSSM (green), E$_6$SSM with $M_{Z^\prime} = 2.5$ TeV (blue) and
  E$_6$SSM with $M_{Z^\prime} = 4.5$ TeV (red).}
\label{Fig:e6ssmvsmssm}
\end{center}
\end{figure}
In Fig \ref{Fig:e6ssmvsmssm} results from the scan are plotted showing
the tuning for each case against the lightest Higgs mass.  As expected
the dependence on the Higgs mass is now quite weak, while the minimum
tuning in the model for the E$_6$SSM is increased by the mass of the
$Z^\prime$ boson.  So in the case of a very low cutoff the tuning
required to get a 125 GeV Higgs is not so large.  However the tuning
from the $Z^\prime$ mass appears already at tree level and is
therefore not heavily suprressed when the cutoff scale is low. We find
that in the E$_6$SSM current limits on the mass of the $Z^\prime$
boson imply that the E$_6$SSM fine tuning is at least 100.
Additionally if run II of the LHC pushes the limit on the $Z^\prime$
mass to be above $4.5$ TeV then the fine tuning in the model will be
greater than at least 350.

This demopnstrates two important points about these $U(1)$
extensions. Firstly that limits on the $Z^\prime$ mass play an
incredibly important role in constraining natural scenarios in such
models and secondly that the tuning from the $Z^\prime$ limits in
these models depends on less assumptions about SUSY breaking than the
tuning required by the $125$ GeV Higgs measurement which concerns
people in the MSSM.

However we should also acknowledge that there are other limits which
play a similar role.  Chargino limits directly constrain the $\mu$
parameter (or effective $\mu$ parameter in these $U(1)$ extensions).
However the chargino limits from the LHC are scenario dependent and
depending on whether there are light sleptons or sneutralinos and the mass
difference between the lightest chargino and lightest neutralino.
Current limits placed by CMS and ATLAS extend up to $m_{\chi^\pm_1}
\approx 700-740$ GeV if there are light
sleptons\cite{Khachatryan:2014qwa,Aad:2014nua} with much weaker bounds
if there are no light sleptons or sneutrinos\footnote{Useful summary
  plots of these limit may be found on the public pages of ATLAS,
  \url{https://atlas.web.cern.ch/Atlas/GROUPS/PHYSICS/CombinedSummaryPlots/SUSY/ATLAS_SUSY_EWSummary/ATLAS_SUSY_EWSummary.png}
  and CMS
  \url{http://cms.web.cern.ch/sites/cms.web.cern.ch/files/styles/large/public/field/image/Image_03_exclusion_Combined.png?itok=8FMBpu_1}.}.

Nonetheless the MSSM the impact of chargino limit is shown in
Fig.~\ref{Fig:chargino-plot}.  There we see that if the the full
  parameter space with $m_{\chi^\pm} < 700$ GeV was excluded the
  impact would be to make the tuning in the MSSM with a $20$ TeV
  cutoff slightly worse than that of the E$_6$SSM with the same cutoff
  and a $Z^\prime$ mass just larger than current limits.  would be
  similar to that can also be attcked more directly by looking at
  chargino mass limits as this rasies $\mu$ directly.  In the
  E$_6$SSM, while raising the chargino limit can have the same impact
  in principle, due to current limits on the $Z^\prime$ mass already
  imposing a significant degree of tuning, chargino masses do not make
  much of a noticable change.

\begin{figure}[h]
\begin{center}
\resizebox{!}{10cm}{%
\includegraphics{mssm_chargino_plot.png}}%%
\caption{Scatter plot of fine tuning vs lightest Higgs mass in the MSSM with $m_{\chi^\pm_1} = 200$ GeV shown in green $m_{\chi^\pm_1} = 500$ GeV in blue and $m_{\chi^\pm_1} = 700$ GeV in red.}
\label{Fig:chargino-plot}
\end{center}
\end{figure}

The exact level of tuning from the $Z^\prime$ depends on the charges
of the extra $U(1)$ gauge symmetry it is associated with.  In
Fig.~\ref{Fig:othere6modelsvsmssm} we look at the fine tuning for
other U(1) extensions for the same $Z^\prime$ masses as we did for the
E$_6$SSM.  To simplify the analysis we fix $\tan \beta = 10$, but scan
over the remaining parameters as in Table \ref{tab:scanranges} and fix
the rest to the same values we did in the scan carried out for
Fig.~\ref{Fig:e6ssmvsmssm}.

\begin{figure}
\begin{center}
\resizebox{!}{5cm}{%
\includegraphics{e6inert_tb10_tuning_vs_higgs.png}}
\resizebox{!}{5cm}{%
\includegraphics{e6psi_tb10_tuning_vs_higgs.png}}\\
\resizebox{!}{5cm}{%
\includegraphics{e6eta_tb10_tuning_vs_higgs.png}}%%
\caption{Scatter plot of fine tuning vs lightest Higgs mass in ... with $M_{Z^\prime} = 2.5$ TeV shown in blue and $M_{Z^\prime} = 4.5$ TeV in red.}
\label{Fig:othere6modelsvsmssm}
\end{center}
\end{figure}

@@

Interestingly the coefficient $d$ is very small (and negative) in the
case of the $U(1)_I$.  This allows a dramatic reduction in the fine
tuning from the $U(1)_I$ symmetry.  This is a result of the $H_u$
charge associated with $U(1)_I$ vansihing, which means that the
D-terms to the lightest Higgs which is predominantly $H_u$ at large
$\tan \beta$ are suppressed, making it difficult to raise the Higgs
mass in the same way as happens in the other models and explaining why
% DH:: I strongly suspect that they are invalid - but running scan
% to check this currently
the heavier Higgs values are less well populated [can't be obtained? -
are we sure those scattered points are valid?]  than in the other
models.  Therefore the fine tuining behaviour in this model is closer
to that of the MSSM, though the LHC can still constrain the most
natural scenarions from just placing tougher limits on the $Z^\prime$
mass .
